---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---

setwd("~/Dropbox (Personal)/coursera/RepData_PeerAssessment1")

## Loading and preprocessing the data
To load the code and preprocess the data, run the following code block:
```{r loading, echo=TRUE}
    library(data.table)
    activity <- read.table(unz("activity.zip", "activity.csv"),header=T, sep=",")
    activity$date <- as.Date(activity$date, format="%Y-%m-%d")
    summary(activity)
```

## What is mean total number of steps taken per day?
The mean total number of steps taken per day is (ignoring NAs):
```{r mean, echo=TRUE}
    steps_per_day <- tapply(activity$steps, activity$date, function(x) sum(x, na.rm=TRUE))
    hist(steps_per_day, breaks=8)

    mean(steps_per_day)
    median(steps_per_day)
```

## What is the average daily activity pattern?
The average daily activity pattern is
```{r pattern, echo=TRUE}
    plot(steps ~ date, data=activity)
```


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
